### 2015-12-07 paper/1-introduction.tex
Review

%? use paragraphs for layout instead of (dense) flowing text ?
%TODO SIZE: approx. 1-2 column, every TODO is approx. one paragraph
\subsection{DEVS}
%TODO give introduction to the field: DEVS is used for discrete event simulation and is a common basis for most other formalisms
The family of DEVS \cite{DEVSbase} formalisms serve as a common basis for most other discrete event formalisms. Of interest in this paper are the 2 key formalisms: Dynamic Structured \cite{DSDEVS} and Parallel \cite{ParallelDEVS} and their implementation. This project uses the Direct Connection \cite{SymbolicFlattening} algorithm, so from a simulation kernel's perspective only Atomic Models exist linked to each other by ports.

\subsection{Parallel computing}
%TODO parallel computing is necessary with more and more complex systems, blablabla (usual explanation)
Parallel execution of a PDEVS simulation can lower overall runtime and increase the bound on the state space, thereby enabling simulation of more complex systems in the same time-frame.
While the shared memory parallelism offered by most modern hardware does not raise the state space bounds, it can reduce the runtime and offers more direct communication and control between entities involved in synchronization compared to distributed simulation. \\

\subsection{Motivation}
%TODO give motivation (adevs is good, but only conservative; exploit new parallelism features of C++11 too)
Adevs \cite{adevs} offers a very fast conservative synchronized shared memory PDEVS simulator, but no optimistic synchronized variant. 
The latter can be significantly faster, especially in simulations where the runtime behaviour of the simulation is hard to predict. \\
The matured parallelism features of C++11 were used in this project with the dual aim of writing standard-compliant (and thus portable) code and without losing access to powerful low-level threading primitives. % may need another pass on this one :(

\subsection{Solution} 
%TODO give our solution, motivating the need for a different kernel, instead of modifying another kernel (like adevs or PythonPDEVS)
The usage of the Direct connection algorithm makes reusing adevs kernel hard. Devs-ex-machina (dxex) is based on PythonPDEVS \cite{PythonPDEVS}(PyPDEVS) but differs in design and implementation.
The core aim of the project is to offer a deterministic simulation kernel where the simulation modeller is shielded as much as possible from the kernel implementation, without sacrificing performance. As in PyPDEVS, a model need be written only once for use in the different simulation kernels (with the exception of a non-trivial lookahead).\\
The tracing framework from PyPDEVS was ported to allow optional verification of simulations. 
\subsection{Time}
The DEVS formalisms have $\mathbb{R}$ as time base, but any implementation has to decide on an enumerable representation of time. In principle any type with well defined operators can be used as template parameter, but from a performance point of view a type fitting in a machine word offers obvious advantages. In practice, dxex defaults to word sized unsigned integer.
Although this significantly reduces the possible range of virtual time, it avoids approximation errors. The notion of $\epsilon$ is trivial for integer, unlike floating point.\\
Our time representation is extended with a causality field with a range equal or greater than the maximum nr of models in a single kernel ($2^{48}$ by default). If A and B are imminent at time t then $t[1]_a < t[1]_b \oplus t[1]_b < t[1]_a$, while $t[0]_a == t[0]_b == t[0]$. This allows a deterministic ordering of concurrent events. % execution not so much as tracing

>>>>

%? use paragraphs for layout instead of (dense) flowing text ?
%TODO SIZE: approx. 1-2 column, every TODO is approx. one paragraph
\subsection{DEVS}
%TODO give introduction to the field: DEVS is used for discrete event simulation and is a common basis for most other formalisms
The family of DEVS~\cite{DEVSbase} formalisms serve as a common basis for most other discrete event formalisms. In this paper, we are in particular interested in Dynamic Structured DEVS~\cite{DSDEVS} and Parallel DEVS~\cite{ParallelDEVS}.
We propose a DEVS implementation (called DEVS-Ex-Machina and referred to as "dxdev" throughout this paper) that is based on the
the Direct Connection~\cite{SymbolicFlattening} algorithm. This algorithm flattens all hierarchically nested models which leads to, from a simulation kernel's perspective, a collection of Atomic Models linked to each other by ports.

\subsection{Parallel computing}
%TODO parallel computing is necessary with more and more complex systems, blablabla (usual explanation)
One way to speed up the PDEVS simulation is by using parallel execution. By its very nature, this can lower the runtime considerably and increase the bound on the state space, thereby enabling simulation of more complex systems in the same time-frame.
While the shared memory parallelism offered by most modern hardware does not raise the state space bounds, it can reduce the runtime and offers more direct communication and control between entities involved in synchronization compared to distributed simulation.

\subsection{Motivation}
%TODO give motivation (adevs is good, but only conservative; exploit new parallelism features of C++11 too)
One of the drawbacks of adevs \cite{adevs}, a well-known PDEVS simulator which is performing well with conservative shared memory synchronization,
is that it lacks support for optimistic synchronization.
Using optimistic synchronization can be significantly faster, especially in simulations where the runtime behaviour of the simulation is hard to predict.
The matured parallelism features of C++11, on the other hand, created new opportunities as well. They allowed to write very efficient code giving access to powerful low-level threading primitives and yet guarantee standard-compliance and portability.
% may need another pass on this one :(

\subsection{Solution} 
%TODO give our solution, motivating the need for a different kernel, instead of modifying another kernel (like adevs or PythonPDEVS)
The usage of the Direct connection algorithm makes reusing existing kernels (such as adevs or PythonPDEVS~\cite{PythonPDEVS}, also known as PyPDEVS, hard. Dxex is based on PythonPDEVS \cite{PythonPDEVS}(PyPDEVS) but differs in design and implementation.
The core aim of the project is to offer a deterministic simulation kernel where the simulation modeller is shielded as much as possible from the kernel implementation, without sacrificing performance. As in PyPDEVS, a model has to be written only once for use in the different simulation kernels.\\
In order to monitor and investigate simulations, dxex uses a tracing framework which is derived from the one used in PyPDEVS.
\subsection{Time}
The theoretical DEVS formalisms all have $\mathbb{R}$ as time base, but any implementation has to decide on an enumerable representation of time. In principle, any type with well-defined operators could be used, but, from a performance point of view, a type fitting in a machine word offers obvious advantages. In practice, dxex defaults to word sized unsigned integer.
Although this significantly reduces the possible range of virtual time, it avoids approximation errors. The notion of $\epsilon$, moreover, is trivial for an integer representation, unlike floating point.\\
In addition, our time representation is enriched with a causality field the range of which is equal or greater than the maximum number of models in a single kernel ($2^{48}$ by default). If $A$ and $B$ are imminent at time $t$, then $t[1]_a < t[1]_b \oplus t[1]_b < t[1]_a$, while $t[0]_a == t[0]_b == t[0]$. This allows a deterministic ordering of concurrent events. % execution not so much as tracing

### 2015-12-07 paper/2-background.tex
Review

\subsection{Conservative Synchronization}
%TODO explain conservative synchronization, and how it uses locking to wait for continuing the simulation
Conservative synchronization is defined by the invariant that no model will advance in time before it has received all input from any influencing model. \\
This requires the concepts of earliest output time (eot) and earliest input time (eit), which define the timespan within which a model can safely advance. \\ The eit of any model is the minimum of all eot values of (directly) influencing models. A model can simulate up to (but not including) eit, then waits until that value is increased. An important disadvantage here is that the influenced-by relation is always defined at model(link) creation, not at runtime. A model that can influence another, but never does, can severely slow down the protocol. 
Deadlock between models that influence each other and end up waiting on each other can be broken/avoided by a variety of means. In this simulator, the CMB \cite{Chandy:1981:ADS:358598.358613} null-message protocol is used. \\
In our implementation null-time is the timestamp a model is guaranteed to have passed in simulation. More precisely, a null message of time t is a guarantee that any output with timestamp t-$\epsilon$ is already sent.\\
Conservative synchronization relies explicitly on information provided by the model creator in the form of a lookahead, a relative timespan during which the model is insensitive to outside events. This lookahead can be non-trivial to calculate, a simulation writer will in general not be able to determine the exact lookahead of models involved in an experiment without having run the experiment.\\ Conservative kernels can operate if there exists a cyclic dependency between them, but at a quite severe performance penalty, as seen in section 4.
\subsection{Optimistic Synchronization}
%TODO explain optimistic synchronization, and how it uses rollbacks to undo causality errors
Optimistic synchronization allows causality errors to occur but recovers using a roll-back mechanism, the most common of which is Timewarp \cite{Jefferson:1985:VT:3916.3988}.
Whenever a kernel receives an event with a timestamp in the kernel's past, the state of the kernel (and all models)reverts to that time.
Because an optimistic kernel does not wait for other kernels, it can achieve a lower runtime than conservative.
The cost of this approach is higher complexity of the code and larger memory usage for state/message saving.
In contrast to conservative, optimistic does not rely on any domain specific information, it is only sensitive to runtime use of connections, not the probability that they might occur. %reword connections ?
\\
If the (runtime) dependency graph contains a cycle, optimistic can suffer a series of cascading reverts. The cost of reverts can be reduced by lazy cancellation and/or lazy re-evaluation \cite{FujimotoBook}.
\subsection{Global Virtual Time}
To avoid exhausting memory in state/event saving, optimistic synchronization relies on the concept of global virtual time\cite{Jefferson:1985:VT:3916.3988}. In optimistic simulations, GVT is defined as the lowest timestamp of any unprocessed event. \\ Intuitively this is the simulation timepoint that is certain to be preserved, corresponding exactly with a sequential simulation.
In conservative, the minimum simulation time of all kernels is the LBTS, or in terms of null messages: the least timestamp of any null message in transit.
The GVT calculation is vital to safely commit unrecoverable transactions such as IO (e.g. tracing), releasing memory, ... .

\subsection{C++11 Parallelism Features}
%TODO explain what is new in C++11 for parallelism and how it contributes to our cause
C++11 offers a wide range of portable synchronization primitives in the Standard Library, whereas in earlier versions one had to resort to non-portable (C) implementations. More importantly, C++11 is the first version of the standard that actually defines a multi-threaded abstract machine memory model in the language.
Our kernels use a wide range of threading primitives and atomic operations. As an example, eot/eit/nulltime are exchanged not as messages, but as reads/writes to atomic fields shared by all kernels. This avoids the otherwise unavoidable latency penalty by mixing simulation messages with synchronization messages. For an in-depth study, see \cite{CPE:CPE3007}.
Most modern compilers support the full standard, allowing the kernels to be portable by default on any standard compliant platform.

>>>>

\subsection{Conservative Synchronization}
%TODO explain conservative synchronization, and how it uses locking to wait for continuing the simulation
Conservative synchronization is defined by the invariant that no model will advance in time before it has received all input from any influencing model. \\
This requires the concepts of earliest output time (eot) and earliest input time (eit), which define the timespan within which a model can safely advance.\\
The eit of any model is the minimum of all eot values of (directly) influencing models. A model can simulate up to (but not including) eit and then waits until that value is increased. \\
An important disadvantage here is that the relation that calculates the dependencies, is always defined at model (link) creation time, and not at runtime. A model that can influence another model theoretically, but never does in real life, can severely slow down the protocol. \\
Deadlocks between models that influence each other and end up waiting on each other, can be broken/avoided by a variety of means. In dxex, the CMB~\cite{Chandy:1981:ADS:358598.358613} null-message protocol is used. \\
In our implementation, null-time is the time stamp a model is guaranteed to have passed in simulation. More precisely, a null message of time $t$ guarantees that any output with time stamp $t-\epsilon$ has already been sent.\\
Conservative synchronization relies explicitly on information provided by the model creator in the form of a lookahead, a relative timespan during which the model is insensitive to outside events. This lookahead can be non-trivial to calculate; therefore, a simulation writer will in general not be able to determine the exact lookahead of the models involved in an experiment without having run the experiment.\\
Conservative kernels can even operate if there are cyclic dependencies involved, but this can come at a severe performance penalty, as seen in section 4.

\subsection{Optimistic Synchronization}
%TODO explain optimistic synchronization, and how it uses rollbacks to undo causality errors
In the case of optimistic synchronization, causality errors can occur but these will be reverted via a roll-back mechanism, the most common of which is Timewarp \cite{Jefferson:1985:VT:3916.3988}.\\
Whenever a kernel receives an event with a time stamp in the kernel's past, the state of the kernel (and all models) is reverted to that time.\\
Because an optimistic kernel does not wait for other kernels, it can run much faster than a conservative kernel.
The cost of this approach, however, is higher complexity of the code and a larger memory usage for the state and message saving.
In contrast to conservative synchronization, optimistic synchronization does not rely on any domain specific information. Therefore, it is only sensitive to the runtime use of connections, not the probability that they might occur. %reword connections ?\\
If the (runtime) dependency graph contains a cycle, optimistic synchronization might lead to a series of cascading reverts. The cost of reverts can be reduced by lazy cancellation and/or lazy re-evaluation~\cite{FujimotoBook}.

\subsection{Global Virtual Time}
To avoid exhausting memory in state/event saving, optimistic synchronization relies on the concept of global virtual time\cite{Jefferson:1985:VT:3916.3988} or GVT. In optimistic simulations, GVT is defined as the lowest time stamp of any unprocessed event.\\
Intuitively, this is the simulation time point that is certain to be preserved, corresponding exactly with a sequential simulation.
In conservative, the minimum simulation time of all kernels is the lower bound of time stamp or LBTS. In terms of null messages this corresponds to the least time stamp of any null message in transit.
The GVT calculation is vital to safely commit unrecoverable transactions such as IO (e.g. tracing), releasing memory, etc.

\subsection{C++11 Parallelism Features}
%TODO explain what is new in C++11 for parallelism and how it contributes to our cause
C++11 offers a wide range of portable synchronization primitives in the Standard Library, whereas in earlier versions one had to resort to non-portable (C) implementations. More importantly, C++11 is the first version of the standard that actually defines a multi-threaded abstract machine memory model in the language.
Our kernels use a wide range of threading primitives and atomic operations. As an example, eot/eit/nulltime are not exchanged as messages, but as reads/writes to atomic fields shared by all kernels. This avoids the otherwise unavoidable latency penalty by mixing simulation messages with synchronization messages. For an in-depth study, we refer to~\cite{CPE:CPE3007}.
Most modern compilers support the full standard, allowing the kernels to be portable by default on any standard compliant platform.

### 2015-12-07 paper/3-features.tex
Review

%TODO explain that we are briefly going over the main features
\subsection{Based on PythonPDEVS}
%TODO say that it is based on PythonPDEVS --done
The simulator is based on PythonPDEVS, and provides the following features: 
\begin{enumerate}
	\item Direct Connection
	\item Dynamic Structured DEVS
	\item Termination function. If specified, a termination function is applied every simulation round to each model to test whether the simulation can terminate. Only available in single-threaded simulation.
	\item State/Message can have any payload type. Different message types can be used together within the same simulation.
	\item Tracing An asynchronous, thread safe and versatile tracing mechanism allows exact verification of the simulation.
	\item Optimistic and Conservative synchronization of PDEVS.
\end{enumerate}
The implementation tries to adhere to the C++ principle that you don't pay performance-wise for what you don't use. For this reason, the support for a termination function for the multi-threaded kernel was abandoned, as it is non-trivial to implement and had a non-negligible impact on the runtime, even when not in use. Another example is the state saving mechanism, which is only used for optimistic parallel simulation and has no performance impact in conservative parallel simulation.\\
The tracing is not comparable to adevs's listener interface. To be usable in optimistic simulation, the tracing of the simulation has to be reversible and only be committed at GVT points. Furthermore, the framework itself has to be thread safe and deterministic so that a simulation will always produce the exact same output.
%TODO Decide whether we should list the functionality that is _not_ implemented
The following features from PyPDEVS are not present
\begin{enumerate}
	\item Activity tracking and relocation
	\item Serialization
	\item Interactive control. 
	\item Distributed simulation
\end{enumerate}
Serialization in this context is the ability to save/load a complete simulation to disk, not the state saving mechanisms required for TimeWarp. %This state saving has no impact in a single threaded or conservative kernel. -> already mentioned above
\\
Model allocation is done by a derivable allocator object which the user can implement to arrange a more ideal (domain-specific) allocation. If this is omitted, a default (non-activity-aware) allocation stripes the models over the simulation kernels.\\
Debugging tools such as a logger and a graph visualizer are included which can track activity with respect to allocation for later study, but not online/dynamic as is possible in PyPDEVS.

%TODO ... and thus has most features from PythonPDEVS, like Dynamic Structure, termination condition, direct connection, ...
%TODO (size: take as much size as you want, depending on how much is shared with PythonPDEVS)
%TODO (maybe: note that models are not compatible)
\subsection{Different Synchronization protocols}
%TODO explain rationale: why different synchronization protocols
Synchronization is required for PDEVS to prevent or recover from causality violations if the simulation is executed in parallel. Preventing causality violations (conservative) typically requires domain-specific information to compensate for performance loss, while recovering from causality errors (optimistic) requires a state to recover to.

\subsubsection{Conservative}
A conservative kernel will determine which kernels it is influenced by. This information is constructed from the incoming connections on all hosted models. The process is only 1 link deep, since an influenced kernel will in turn be blocked by others deeper in the graph.\\
A model should provide a lookahead function which returns, relative from the current time, the timespan during which the model cannot change state due to an external event. This information is collected for all models hosted on the kernel, and the minimum is set as the lookahead of that kernel. \\
The kernel will calculate its earliest output time and write this value in shared memory. The eit of the kernel is then set as the minimal eot of all influencing kernels. \\
For garbage collection (of sent messages) the LBTS/GVT is calculated as $\min_{\forall i \in \textup{influencors}}( \textup{nulltime}[i])  - \epsilon $.\\
%This is one of the operations that can benefit from using a relaxed memory ordering. % Using it for read atm, check before finalizing and do we need to mention this ?. %writing this here seems inappropriate

\subsubsection{Optimistic}
The optimistic kernel requires from the hosted model only that copying the state is well-defined, which is provided in the base State class for the user. 
The kernels use Mattern's \cite{mattern}
GVT algorithm with a maximum of 2 rounds per iteration to determine a GVT. This process runs asynchronously from the simulation itself. Once found, the controlling thread informs all kernels of the new value, which they can use to execute garbage collection of old states and (anti)messages.

%TODO explain core infrastructure a bit, maybe with a diagram or so, which clearly shows that everything is shared, except the "synchronization" part
%TODO claim that this allows you to easily switch between both methods, to always chose the best one
The user need only provide one implementation of a model for use with both synchronization protocols. A lookahead function is desired to accelerate conservative, but is not required. In the absence of a user supplied lookahead, the kernel assumes it cannot predict beyond its current time + $\epsilon$, creating a lockstep simulation. % implementation detail ?
\\
The implementation details such as defining the copy semantics of a State are provided (but can be overridden). \\
From the user's perspective, the multi-threaded aspect of the kernel is not exposed. % payload protected ?
%TODO indicate in which optimistic is good (when runtime behavior is very hard to predict). Describe some drawbacks (many reverts when everybody influences everybody all the time)

\subsection{Performance Improvements}
%TODO briefly discuss some of the additional changes you implemented
%TODO such as memory pools, memory allocators, profiling-guided-optimizations, ...
Continuous profiling of the kernels in several benchmarks highlighted the following key bottlenecks: 
\subsubsection{Heap}
Events are passed by pointer, avoiding a possibly expensive copy of the payload at the cost of allocation overhead.\\
Allocation becomes prohibitively expensive in highly connected models, so to reduce that overhead we use thread\_local memory pools for states and events, and optionally replace the system malloc with calls to tcmalloc\cite{tcmalloc}. In this way allocating threads do not block each other. If desired, arena-pools are available for single-threaded simulation. A disadvantage is increased complexity in ownership semantics. The creating kernel is responsible for destruction, but this can only be guaranteed up to GVT. Experiments with synchronized pools proved slower than malloc/free. \\
The kernels initially used strings as identifiers, as is done in PyPDEVS. Profiling quickly indicated this to be a performance bottleneck. C++ strings are heap allocated variable sized objects with an atomic reference count, not immutable objects as in Python. Access of that reference count across threads is expensive, as are the calls to malloc/free the string implementation makes to create/destroy new objects, or copy existing.\\
Strings are more intuitive to work with from a user's standpoint, so as a compromise the user can reference models/ports by string name. Once simulation starts all objects use integral identifiers for performance. This also increased usage of the constexpr feature of C++11 in, amongst others, timestamps and message headers.
\subsubsection{Raw pointers}
While an important C++11 feature in general, our initial usage of smart pointers for some types of objects was misplaced. Used across threads, the reference counting becomes very expensive, and the (de)allocating of memory caused significant contention between threads. Models are still held by a smart pointer, as is a kernel, but a message is a raw pointer to compacted memory. 
\subsubsection{Locking}
Locking between kernels uses mostly atomic operations, where we can occasionally leverage memory orderings to only pay for synchronization when we need it. Messages are exchanged via a shared set of queues each with a dedicated lock.\\
On a higher level, we avoid the sending of synchronization messages entirely by writing the timestamp directly into shared memory.\\ Sending of antimessages is fairly cheap in our implementation, since only the modified pointer to the original message is sent to the receiving kernel.
\subsubsection{Schedulers}
PyPDEVS has a wide range of schedulers for the user to choose from, with performance of each depending on the simulation type. Profiling showed in our case that, for a c++ implementation, the heap implementation used in adevs was faster than any of the schedulers we had tested before. Unlike most node based heaps, this scheduler uses a fixed size array where a heap is rebuilt or modified in place depending on the amount of items to update. Items are only updated, never removed.

>>>>

%TODO explain that we are briefly going over the main features
\subsection{Based on PyPDEVS}
%TODO say that it is based on PythonPDEVS --done
Dxex is based on PyPDEVS, and provides the following features: 
\begin{enumerate}
	\item Direct Connection
	\item Dynamic Structured DEVS
	\item Termination function: if specified, a termination function is applied to each model every simulation round to test whether the simulation can terminate. This feature is only available in single-threaded simulations.
	\item The State/Message objects can have any payload type. Different allows different message types to be used within the same simulation.
	\item Tracing: An asynchronous, thread safe and versatile tracing mechanism allows exact verification of the simulation.
	\item Optimistic and Conservative synchronization
\end{enumerate}
Furthermore, the implementation tries to adhere to the C++ principle that you don't pay performance-wise for what you don't use. For this reason, the support for a termination function for the multi-threaded kernel was abandoned, as it is non-trivial to implement and had an adverse impact on the runtime, even when not in use. Another example is the state saving mechanism, which is only used for optimistic parallel simulation and has no performance impact in a conservative parallel simulation.\\
Our tracing implementation is not comparable to adevs's listener interface. To be usable in an optimistic simulation, the tracing of the simulation has to be reversible and only be committed at GVT points. Furthermore, the framework itself has to be thread safe and deterministic so that a simulation will always produce the exact same output.\\
The following features from PyPDEVS are not present
\begin{enumerate}
	\item Activity tracking and relocation
	\item Serialization: in this context this is the ability to save/load a complete simulation to disk, not the state saving mechanisms required for TimeWarp. 
	\item Interactive control
	\item Distributed simulation
\end{enumerate}
In dxex, the model allocation is realized by a derivable allocator object which the user can implement to arrange a more ideal (domain-specific) allocation. If this is omitted, a default (non-activity-aware) allocation stripes the models over the simulation kernels.\\
In addition, debugging tools such as a logger and a graph visualizer are included that can track activity with respect to the allocation for later inspection (but not are not online and as dynamic as is in PyPDEVS).\\
%TODO ... and thus has most features from PythonPDEVS, like Dynamic Structure, termination condition, direct connection, ...
%TODO (size: take as much size as you want, depending on how much is shared with PythonPDEVS)
%TODO (maybe: note that models are not compatible)
\subsection{Different Synchronization protocols}
%TODO explain rationale: why different synchronization protocols
For parallel executions, synchronization is required for PDEVS to prevent causality violations from happening or to recover from causality violations. Preventing causality violations (conservative) typically requires domain-specific information to compensate for the performance loss, while recovering from causality errors (optimistic) requires the calculation of a state to revert to.

\subsubsection{Conservative}
In case of conservative synchronization, any kernel will determine which kernels it is influenced by. This information is constructed from the incoming connections on all hosted models. The process is only 1 link deep, since an influenced kernel will in turn be blocked by others deeper in the graph.\\
A model should provide a lookahead function which returns, relative to the current time, the timespan during which the model cannot change state due to an external event. This information is collected for all models hosted on the kernel and the minimum time is set as the lookahead of that kernel. \\
The kernel will calculate its earliest output time and write this value in shared memory. The eit of the kernel is then set as the minimal eot of all influencing kernels. \\
For the garbage collection (of sent messages), the LBTS/GVT is calculated as $\min_{\forall i \in \textup{influencors}}( \textup{nulltime}[i])  - \epsilon $.\\
%This is one of the operations that can benefit from using a relaxed memory ordering. % Using it for read atm, check before finalizing and do we need to mention this ?. %writing this here seems inappropriate

\subsubsection{Optimistic}
The optimistic kernel requires from the hosted model that copying the state is done carefully (avoid unnecessary copies and make a copies whenever necessary).\\
The kernels use Mattern's \cite{mattern}
GVT algorithm with a maximum of 2 rounds per iteration to determine a GVT. This process runs asynchronously from the simulation itself. Once found, the controlling thread informs all kernels of the new value, which they can use to execute garbage collection of old states and (anti)messages.\\
%TODO explain core infrastructure a bit, maybe with a diagram or so, which clearly shows that everything is shared, except the "synchronization" part
%TODO claim that this allows you to easily switch between both methods, to always chose the best one
The user must only provide one implementation of a model that can be used for both synchronization protocols. A lookahead function is desired to accelerate the conservative protocol, but is not required. In the absence of a user supplied lookahead, the kernel assumes it cannot predict beyond its current time $t+\epsilon$, creating a lockstep simulation. % implementation detail ?\\
The user is shielded from the multi-threaded aspect of the kernel.\\
From the user's perspective, the multi-threaded aspect of the kernel is not exposed. % payload protected ?
%TODO indicate in which optimistic is good (when runtime behavior is very hard to predict). Describe some drawbacks (many reverts when everybody influences everybody all the time)

\subsection{Performance Improvements}
%TODO briefly discuss some of the additional changes you implemented
%TODO such as memory pools, memory allocators, profiling-guided-optimizations, ...
We now discuss a number of bottlenecks that were discovered during the profiling of several benchmarks.
\subsubsection{Heap}
In dxex, events are always passed with pointers and thus involve a possibly expensive copy of the payload caused by allocation and copying overhead. In highly connected models, this allocation cost can become prohibitively expensive, so to reduce that overhead we use thread\_local memory pools for states and events, and, optionally, replace the system malloc with calls to tcmalloc~\cite{tcmalloc}. In this way, allocating threads do not block each other. If desired, arena-pools are available for single-threaded simulation. A disadvantage is increased complexity in ownership semantics. The creating kernel is responsible for destruction, but this can only be guaranteed up to the GVT. Experiments with synchronized pools proved to be slower than implementations with the standard malloc/free.\\
Initially, dxes was using strings as identifiers as was also done in PyPDEVS. Profiling quickly indicated that this caused a real performance bottleneck. In C++, strings are heap allocated variable sized objects with an atomic reference count and not not immutable objects as in Python. Access of that reference count across threads turned out to be quite expensive, as are the calls to malloc/free the string implementation makes to create/destroy new objects or copy existing objects.\\
Strings, however, are more intuitive to work with from a user's perspective, so, as a compromise, we allowed the user to reference models/ports by string name. Once the simulation starts however, all objects use integral identifiers. This also increased the usage of the constexpr feature of C++11 in, amongst others, time stamps and message headers.
\subsubsection{Raw pointers}
While an important C++11 feature in general, our initial usage of smart pointers for some types of objects was misplaced. Used across threads, the reference counting became very expensive, and the (de)allocation of memory caused significant contention between threads. So for the models and the kernels, dxex is still using smart pointers whereas for the messeages, a raw pointer to compacted memory is being used. 
\subsubsection{Locking}
Locking between kernels uses mostly atomic operations and, occasionally, we can leverage memory orderings to only pay for synchronization when we need it. Messages, on the other hand, are exchanged via a shared set of queues each with a dedicated lock.\\
On a higher level, we avoid the sending of synchronization messages entirely by writing the time stamp directly into shared memory.\\ Sending of antimessages is fairly cheap in our implementation, since only the modified pointer to the original message is sent to the receiving kernel.
\subsubsection{Schedulers}
PyPDEVS has a wide range of schedulers to choose from with varying performance depending on the simulation type. Profiling showed that, the heap implementation used in adevs was faster than any of the schedulers we had tested before (in a C++ environment). Unlike most node based heaps, this scheduler uses a fixed size array where its heap is rebuilt or modified in place depending on the amount of items to update. Items are only updated, never removed.
